---
title: "Tidymodels and Vetiver"
format: html
editor_options: 
  chunk_output_type: console
  canonical: true
---

## Goals

The goals of this activity are to:

-   Create two models using the `tidymodels` framework that predict if a Seattle Ferry will depart on time or be delayed from a specific station.
-   Save a model as a pin to Posit Connect using the `vetiver` package.
-   Deploy a model to Posit Connect as an API using the `vetiver` package.
-   Learn how to version and monitor your machine-learning models using `vetiver`.

## Overview

1.  Import validated ferry-weather dataset from the database.
2.  Transform the data for modeling.
3.  Create `logistic_reg()` and `xgboost()` classification models using `tidymodels`.
4.  Assess the performance of both models using `tidymodels`.
5.  Save each model as a pin to Posit Connect using the `vetiver` package.
6.  Use the pinned models to serve a plumber API hosted on Posit Connect.
7.  Use the `vetiver` and `pins` package to version and monitor the performance of the models over time.

## Setup

```{r}
#| label: setup

library(tidyverse)
library(pins)
library(tidymodels)
library(lubridate)
library(skimr)
library(vetiver)
library(xgboost)
library(plumber)
library(rsconnect)

```

## Task 1 - Create a tidymodel

[Tidymodels](https://www.tidymodels.org/) is a collection of R packages for modeling and machine learning using [tidyverse](https://www.tidyverse.org/) principles. While this workshop is **not** a modeling workshop, to demonstrate the utility of the `vetiver`, we need to create a model.

For this activity, we will train and fit a logistic regression model to predict whether a Seattle Ferry departs a station on time or is delayed. Later on, we will create another model using xgboost.

🔄 Tasks

-   Explore the `ferry-weather` dataset. Use any method you'd like to understand the data. Try functions like `glimpse()`, `str()`, and `View()`.

-   Create a logistic regression tidymodel that predicts `delay_status`.

#### Read and Explore the `ferry-weather` dataset

```{r}
#| label: read-data

# Read in dataset from database
#!#!#!#!#!#!#!#!#! Need to update with database, not pin
board <- board_connect(auth = "manual", 
                       server = "https://connect.posit.it", 
                       key = Sys.getenv("CONNECT_POS_IT_API_KEY"))
ferry_weather <- pin_read(board, "katie.masiello/modeldata_validated") |> 
  as_tibble()
```

```{r}
#| label: explore-data

# Use the space below to explore the ferry-weather dataset
# Try functions like glimpse(), str(), skim(), and View()

glimpse(ferry_weather)
str(ferry_weather)
skim(ferry_weather)
View(ferry_weather)

```

#### Prep Data for Modeling

Sometimes, you may need to transform the original dataset to make it conducive to modeling. In the below code cell, read the comment above each bit of code to understand what types of transformations were conducted.

```{r}
#| label: prep-data

# DO NOT EDIT THIS CODE CELL, JUST RUN IT

ferry_weather_prepped <- ferry_weather |> 
  
  # Select for columns of interest
  #  Removed vessel but kept departing, since this would cause a rank
  #  deficient fit (all vessels don't depart from all places)
  select(departing, 
         closest_hour, 
         date, 
         weather_code, 
         wind_speed_10m, 
         wind_gusts_10m, 
         delay) |> 
  
  # Convert delay to classification
  #  If more than 3 minutes late --> "delayed"
  mutate(delay_status = case_when(
    delay <= 3 ~ "on-time",
    delay > 3 ~ "delayed" 
  )) |> 
  
  # Convert wind to classification
  #  If more than 10mph --> "windy"
  mutate(wind_status = case_when(
    wind_speed_10m >= 10 ~ "windy",
    wind_speed_10m < 10 ~ "calm"
  )) |> 
  select(-wind_speed_10m) |> 
  
  # Convert gust to classification
  #  If more than 15mph --> "gusty"
  mutate(gust_status = case_when(
    wind_gusts_10m >= 15 ~ "gusty",
    wind_gusts_10m < 15 ~ "calm"
  )) |> 
  select(-wind_gusts_10m) |>
  
  # Convert dates to ymd
  mutate(date = ymd(date)) |>
  
  # Extract closest hour (just the hour)
  mutate(closest_hour = ymd_hms(closest_hour, truncated = 3)) |> 
  mutate(hour = as.factor(hour(closest_hour))) |> 
  select(-closest_hour) |> 
  
  # Change weather code to factor
  mutate(weather_code = as.factor(weather_code)) |> 
  
  # Convert chr to factors
  mutate_if(is.character, as.factor) |> 
  
  # Exclude missing data
  na.omit() |> 
  
  # Clean departing column
  mutate(departing = str_to_title(str_replace_all(departing, "_", " ")))

```

Feel free to explore the prepped dataset!

#### Split Data

For modeling, you often split your data into a **training and testing** dataset. Before we do that, we'll first extract the last 500 rows, which will be used later on for model monitoring.

```{r}
#| label: split-data

# Monitoring data (will be used later on)
monitoring_data <- tail(ferry_weather_prepped, 500)

# Put 3/4 of the data into the training set 
data_split <- initial_split(head(ferry_weather_prepped, -500), prop = 3/4)

# Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```

#### Create the Recipe

```{r}
#| label: create-recipe

ferry_rec <- 
  
  # Add recipe (what does the model do?)
  recipe(delay_status ~ ., data = train_data) |> 
  
  # Remove delay from modeling
  step_rm(delay) |> 
  
  # Extract predictors like day of week and month from date
  step_date(date, features = c("dow", "month"), keep_original_cols = FALSE) |> 
  
  # Convert all factors to binary "dummy variables"
  step_dummy(all_nominal_predictors())
```

#### Define Model Type

We'll first use logistic regression via *generalized linear model*, or glm. This method is used for predicting binary outcomes (delayed vs on-time).

```{r}
#| label: define-model-type

# Specify model using parsnip
ferry_mod_lr <- 
  logistic_reg() |> 
  set_engine("glm")

```

#### Create the Workflow

```{r}
#| label: define-workflow

# Bundle parsnip model with recipe
ferry_workflow_lr <- 
  workflow() |> 
  add_model(ferry_mod_lr) |> 
  add_recipe(ferry_rec)
```

### Train the Model

Train the model using our training set. This step may take a few seconds.

```{r}
#| label: train-model

# Train the model
ferry_fit_lr <- 
  ferry_workflow_lr |> 
  fit(data = train_data)
```

#### Evaluate the Model

We'll use `tidymodels` to evaluate how our model did using the testing dataset. Remember that this model is **far from perfect** and we don't care too much about making a perfect model.

```{r}
#| label: evaluate-model

# Make Prediction
predictions_lr <- augment(ferry_fit_lr, test_data)

# Evaluate the model
metrics_lr <- predictions_lr %>%
  metrics(truth = delay_status, estimate = .pred_class)

metrics_lr

# Confusion Matrix
predictions_lr |> 
  conf_mat(delay_status, .pred_class)

```

🛑 Stop here!

## Task 2 - Pin Model to Posit Connect

Now that we have a model, we'll want to save it to a location so that it can be easily shared with colleagues, other content, or your future self! The `pins` package is the workhouse behind this workflow. The `vetiver` package has some wrappers around the pinning functions that we'll use in this task.

#### Create a Vetiver Model

Before we go any further, we need to convert our logistic regression model into a format that can be used by `vetiver`:

```{r}
#| label: create-vetiver-model

# Create Vetiver Model
v_model <- vetiver_model(ferry_fit_lr, "ferry_model")

```

And now we can save it, or *pin it*, to Posit Connect! Remember to define your pinning board as Posit Connect.

```{r}
#| label: pin-model

#!#!#!#!#!#! modify so it uses ferryland Posit Connect

# Register Posit Connect as board
# board <- board_connect(auth = "manual", 
#                        server = "https://connect.posit.it", 
#                        key = Sys.getenv("CONNECT_POS_IT_API_KEY"))

# Write Vetiver Model
board |> vetiver_pin_write(v_model)

```

Navigate to Posit Connect and check out your pinned model!

🛑 Stop here!

## Task 3 - Serve Model as an API

APIs are a great way for users, applications, or other systems to interact with your model. `vetiver` leverages [Plumber](https://www.rplumber.io/) which is used to create APIs with only R code! Let's create an API here within Posit Workbench, and then deploy it to Posit Connect.

**Be sure to add your user name for the pin!**

```{r}
#| label: create-api

# Create API from pinned model using vetiver
#  Swap out "ryan" for your username on Connect.
vetiver_deploy_rsconnect(board = board, name = "ryan/ferry_model")

```

Once it's deployed, navigate to the API on Posit Connect and give it a test run! Click the operation on the left called `Return predictions from model using 7 features`. Then copy any variation of the request below and click **Try**. You should see a response of either `delayed` or `on-time`.

```         
[
  {
    "departing": "Anacortes",
    "date": "2024-07-05",
    "weather_code": "3",
    "delay": 0,
    "wind_status": "windy",
    "gust_status": "calm",
    "hour": "10"
  }
]
```

#### Make a Prediction using R Code

On Posit Connect, navigate to your newly created API, and copy the URL for the content below. You can find this URL by clicking the access tab --\> URL.

```{r}
api_url <- "https://connect.posit.it/content/0b0a63a0-ec5b-4ecb-bee3-4e7256249981"

# Append "/predict" to the end of your api to create the endpoint
endpoint <- vetiver_endpoint(paste0(api_url, "/predict"))

endpoint
```

Below is a data frame representing information on a ferry departing. Let's see if it is predicted to be delayed or ontime. Feel free to modify.

```{r}
#| label: make-new-ferry-data

# Create new ferry data point
new_ferry <- tibble(
  departing = "Friday Harbor",
  date = "2024-07-12",
  weather_code = "3",
  delay = 0,# This feature is not considered by the model
  wind_status = "calm",
  gust_status = "calm",
  hour = "6"
)

```

And now we can use the `predict()` function to query the API so that it returns a prediction of delayed or on-time. Note that we have to add our API key to the predict function since this API is only accessible to specific users!

```{r}
#| label: predict-r

# Predict from Vetiver endpoint
predict(endpoint, 
        new_ferry,
        httr::add_headers(Authorization = paste("Key", Sys.getenv("CONNECT_POS_IT_API_KEY"))))$.pred_class
```

If you'd like to learn more about creating API keys on Posit Connect, check out the link [here](https://docs.posit.co/connect/user/vetiver/#predict-from-your-model-endpoint).

🛑 Stop here!
