---
title: "Model Card: Ferry Model - R Focus"
date: '`r Sys.Date()`'
output: 
  html_document
params:
    board: !r pins::board_connect()
    name: ryjohnson09/ferry_model
    version: NULL
---

```{r setup, include=FALSE}
library(tidyverse)
library(vetiver)
library(pins)
library(yardstick)
knitr::opts_chunk$set(echo = FALSE)
v <- vetiver_pin_read(params$board, params$name, version = params$version)
v_meta <- pin_meta(params$board, params$name)
theme_set(theme_light())
```

A [model card](https://doi.org/10.1145/3287560.3287596) provides brief, transparent, responsible reporting for a trained machine learning model.

## Model details

-   Developed by Ryan Johnson
-   `r cli::pluralize("{v$description} using {ncol(v$prototype)} feature{?s}")`
-   The model predicts if a Seattle Ferry will depart a station late or on time. Departures greater than 3 minutes of expected departure time are considered `delayed`.
-   The model leverages time, day, and weather to predict if a ferry is `delayed` or `on-time`.
-   Version `r v$metadata$version` of this model was published at `r v_meta$created`
-   Citation and/or license details for the model
-   If you have questions about this model, please contact `ryan@posit.co`.

## Intended use

-   The primary intended use of this model is for educational purposes only for the 2024 Posit Conference Data Science Workflows Workshop.
-   The primary intended users of this model are participants of the workshop.
-   This model was designed for simplicity and to teach the basics of `tidymodels`. It does not predict well and should not be used for actual ferry delay predictions.

## Metrics

-   The metrics used to evaluate this model are `accuracy`, `kap`, and a confusion matrix.

## Training data

```{r}
glimpse(v$prototype)
```

## Caveats & recommendations

-   To reiterate, this model is for educational purposes only!
