---
title: Using `{connectapi}` 
editor: visual
editor_options: 
  chunk_output_type: console
  canonical: true
---

## Goals

The goals of this activity are to:

Use the `{connectapi}` package to:

-   Retrieve historical content usage data
-   Add your neighbor as a viewer to your content item  

## Setup

```{r}
#| label: setup

library(connectapi)
library(tidyverse)

```

## Task 1 - Retrieving data on historical content views

🔄 Task

First let's see what content YOU have published to Connect. 

```{r}
#| label: get my content information

# Make a connection to the Connect server
client <- connect()

# Determine your unique "user_guid" to filter content that you own
my_user_guid <- client$GET("/v1/user")$guid

# Retrieve all content that you own 
content_df <- get_content(client, owner_guid = my_user_guid)

# Let's look at it
content_df 

# Or a narrower data frame with columns of interest
content_df |> 
    select(guid, name, title, content_url, dashboard_url)
```

We'll get usage information for a content item of interest. As a suggestion, how about choosing your model API because it has gotten some use. 

Important Note - In the Workshop, your API key is scoped as a "Publisher" so the results below will be limited to content that you own. You will not be able to pull usage data for content that you do not own.

```{r}
#| label: get the content guid for the model API
 
content_guid <- content_df |> 
    filter(str_detect(name, "ferry_model")) |> 
    pull(guid)

```

```{r}
#| label: get usage data for the content item

# At the moment, there are two different endpoints for usage data, depending on if the content is Shiny (R/Python) or not. We'll determine which endpoint to use based on the `app_mode` field in the content information. If you are using the guid for your API, the type will be "api" (aka "non-shiny")

type <- content_df |> 
    filter(guid == content_guid) |> 
    pull(app_mode)

# Query appropriate telemetry endpoint for usage data
if (type == "shiny" || type == "python-shiny") {
  usage <- get_usage_shiny(client,
    content_guid = content_guid,
    limit = Inf
  ) 
} else {
  usage <- get_usage_static(client,
    content_guid = content_guid,
    limit = Inf
  ) 
}

usage

```


## Task 2 - Add your neighbor as a viewer to your content

# For the specified content item add the desired user
content <- content_item(client, content_guid)
content_add_user(content, user_guid, role = access_type)

🔄 Task

Now let's use the API to add someone as a viewer to your content item.

```{r}
#| label: get user information

# Get user details
# Notice that with your "Publisher" API key, some of the user information is not visible to you 
all_users <- get_users(client, limit = Inf) |> select(username, first_name, last_name, guid)

my_neighbor_username <- "________"

user_guid <- all_users |> 
    filter(username == my_neighbor_username) |> 
    pull(guid)

access_type <- "viewer"

# Add your neighbor as a viewer to the content item
content <- content_item(client, content_guid)
content_add_user(content, user_guid, role = access_type)
```

Go to your content URL and verify the results!